---
title: "Machine Learning Course Project: Human Activity Recognition"
author: "Nathaniel Reed"
date: "December 24, 2015"
output: html_document
---

```{r, results='hide', message=FALSE, echo=FALSE, warning=FALSE}
library(caret)
library(dplyr)
library(lattice)
library(gridExtra)
library(randomForest)
library(scatterplot3d)
```

## Introduction

In this brief report, we use the Human Activity Recognition data set to analyze the movement patterns of test subjects and fit a predictive model that accurately classifies the type of movement given a training and test sets that contain sensor data. We use Random Forest and analyze the important variables and the error rate on the out-of-bag samples used by this algorithm, then show the predicted values.

## Getting and Cleaning the Data

```{r}
training <- read.csv("pml-training.csv")
````

Subset by removing variables with missing data. In addition, to avoid over-fitting, we remove the timestamps and index variable "X":

```{r}
toDrop <- c("kurtosis", "skewness", "avg_", "stddev", "var", "amplitude", "min", "max", "cvtd_timestamp", "num_window", "new_window", "user_name", "X", "raw_timestamp")
training <- subset(training, select = -grep(paste(toDrop, collapse="|"), names(training)))
```

## Exploratory Analysis: Feature plots

Density plots show the distributions of variables for each "classe" factor. There is a distinctive pattern per variable for each class of movement. Notice the arm and belt measurements below:

```{r}
featurePlot(x=training[,c("roll_belt", "pitch_belt", "yaw_belt")], y=training$classe, plot="density", auto.key=list(columns=5))
featurePlot(x=training[,c("gyros_arm_x", "gyros_arm_y", "gyros_arm_z")], y=training$classe, plot="density", auto.key=list(columns=5))
```

The extents and averages of each variable can also tell us a lot about the movement. The minimum, maximum and mean measurements for the gyroscope x, y and z measurements are shown below. There seems to be a strong relationship between each variable and the type of movement:

```{r}
by_classe <- group_by(training, classe) %>% mutate(min_gyros_arm_x = min(gyros_arm_x),
                                               min_gyros_arm_y = min(gyros_arm_y),
                                               min_gyros_arm_z = min(gyros_arm_z),
                                               max_gyros_arm_x = max(gyros_arm_x),
                                               max_gyros_arm_y = max(gyros_arm_y),
                                               max_gyros_arm_z = max(gyros_arm_z),
                                               mean_gyros_arm_x = mean(gyros_arm_x),
                                               mean_gyros_arm_y = mean(gyros_arm_y),
                                               mean_gyros_arm_z = mean(gyros_arm_z)) 
p1 <- xyplot(by_classe$min_gyros_arm_x ~ by_classe$classe, fill.color=colors, groups=by_classe$classe, pch=c(1,2,3,4,5), xlab = "Movement type", ylab="Min X")
p2 <- xyplot(by_classe$min_gyros_arm_y ~ by_classe$classe, fill.color=colors, groups=by_classe$classe, pch=c(1,2,3,4,5), xlab = "Movement type", ylab="Min Y")
p3 <- xyplot(by_classe$min_gyros_arm_z ~ by_classe$classe, fill.color=colors, groups=by_classe$classe, pch=c(1,2,3,4,5), xlab = "Movement type", ylab="Min Z")

p4 <- xyplot(by_classe$max_gyros_arm_x ~ by_classe$classe, fill.color=colors, groups=by_classe$classe, pch=c(1,2,3,4,5), xlab = "Movement type", ylab="Max X")
p5 <- xyplot(by_classe$max_gyros_arm_y ~ by_classe$classe, fill.color=colors, groups=by_classe$classe, pch=c(1,2,3,4,5), xlab = "Movement type", ylab="Max Y")
p6 <- xyplot(by_classe$max_gyros_arm_z ~ by_classe$classe, fill.color=colors, groups=by_classe$classe, pch=c(1,2,3,4,5), xlab = "Movement type", ylab="Max Z")

p7 <- xyplot(by_classe$mean_gyros_arm_x ~ by_classe$classe, fill.color=colors, groups=by_classe$classe, pch=c(1,2,3,4,5), xlab = "Movement type", ylab="Mean X")
p8 <- xyplot(by_classe$mean_gyros_arm_y ~ by_classe$classe, fill.color=colors, groups=by_classe$classe, pch=c(1,2,3,4,5), xlab = "Movement type", ylab="Mean Y")
p9 <- xyplot(by_classe$mean_gyros_arm_z ~ by_classe$classe, fill.color=colors, groups=by_classe$classe, pch=c(1,2,3,4,5), xlab = "Movement type", ylab="Mean Z")

grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,ncol=3)
```

## Partition the Data

To speed up the building of our model and to validate the model we will use a portion of the data as a training set and the remainder for validation:

```{r}
InTrain<-createDataPartition(y=training$classe,p=0.3,list=FALSE)
training_subset<-training[InTrain,]
testing_subset<-training[-InTrain,]
```

## Model

For performance, we will register multiple cores with the multicore package, which will allow train() to build multiple trees in parallel. Your results may vary, depending on your hardware:

```{r, message=FALSE, warning=FALSE}
library(doMC)
registerDoMC(5)
```

Next we build the model using random forests with with 5-fold cross-validation, setting allowParallel to enable train to take advantage of parallel processing:

```{r}
modFit <- train(classe~.,data=training_subset,method="rf",
                trControl=trainControl(method="cv",number=5),
                prox=TRUE,allowParallel=TRUE)
varImp(modFit)
```

## Cross-Validation

The resulting model is highly accurate, with an error rate of approximately 2%. We would like to see how accurate it is on the hold-out set of testing data, too:

```{r}
predictions <- predict(modFit, testing_subset)
confusionMatrix(predictions, testing_subset$classe)
```

We see that the model performs well on the testing data, too, with an accuracy of 98%.

## Analyzing Model Fit

It's difficult to visualize the centers of the clusters in a data set with a high number of dimensions like this one, but we can look at some of the most important variables. The belt measurements for roll, picth and yaw are ranked most important in the random forest model we built. Here we can see the cluster centers in 3 dimensions (we could use a similar approach for the x, y and z of the other sensors):

```{r, echo=TRUE, warning=FALSE}
classeP <- classCenter(training_subset[,c("roll_belt", "pitch_belt", "yaw_belt")], training_subset$classe, modFit$finalModel$proximity)
classeP <- as.data.frame(classeP); classeP$classe <- rownames(classeP)
colors <- c("red", "blue", "yellow", "green", "purple")
colors <- colors[as.numeric(training_subset$classe)]
s3d <- scatterplot3d(x=training_subset$roll_belt, y=training_subset$pitch_belt, z=training_subset$yaw_belt, type="h", color=colors) 
s3d$points3d(x=classeP$roll_belt, 
             y=classeP$pitch_belt, 
             z=classeP$yaw_belt, 
             type="h", data=classeP, pch=c(1,2,3,4,5), col="black") 
```

## Source Code

The source code (.Rmd and .html) files for this page are available at http://github.com/natereed/coursera-machine-learning.
